---
title: "Cancer incidence and HWSE path analysis"
subtitle: "GM-UAW Cohort Study"
# author: "Kevin Chen"
date: \today
fontsize: 11pt
output:
  pdf_document:
    # latex_engine: lualatex
    includes:
      in_header: "~/HeadRs/StatHead.sty"
    number_sections: false
    toc: false
  keep_tex: true
  # word_document:
  #   reference_docx: "~/HeadRs/style-guide.docx"
  # beamer_presentation:
  #   toc: false
  #   slide_level: 2
  #   # latex_engine: lualatex
  #   includes:
  #     in_header: "~/HeadRs/BeamerHead.sty"
bibliography: "`r path.expand('~/eisen/gm/references/gm.bib')`"
csl: "`r path.expand('~/HeadRs/csl/ama.csl')`"
geometry: margin=2.54cm
---

\onehalfspacing
<!-- \renewcommand{\arraystretch}{1.1} -->
\fancyhead[L]{Cancer incidence (including SEER data)}
\renewcommand{\headrulewidth}{0pt}

This packet summarizes Cox proportional hazard model results relating cumulative exposure to straight, soluble, and synthetic metalworking fluids (MWF) to cancer incidence of 14 types and for all sites combined (includes types not shown here) in the UAW-GM Cohort. The types are colon, rectal, pancreatic, esophageal, stomach, laryngeal, lung & bronchial, breast, prostate, kidney and renal pelvic, bladder, melanoma, leukemia, and non-Hodgkin lymphoma. In addition, path analyses testing for the presence of healthy worker survivor effect ar presented. These include Cox models relating emloyment status to cancer incidence and one Cox model relating cumulative MWF exposure to leaving work. The path analyses largely follow the methods outlined in Garcia et al.[@Garcia_2017]

The study population, summarized in Table \ref{tab:tab1}, includes subjects from the UAW-GM Cohort who were still alive at the start of cancer incidence follow-up. Cancer incidence data were abstracted from Michigan Cancer Registry for all plants from years 1985 onward. Cancer incidence data for plants 1 and 2 were supplemented with data from the Surveillance, Epidemiology, and End Results Program (SEER). Follow-up begins three years after hire and no earlier than 1973 for plants 1 and 2 or 1985 for plant 3. Follow-up ends upon reaching the the oldest observed age at death (considered lost to follow-up), death, cancer incidence, or the year 2015, whichever comes first. In the path analyses for employment status and cancer incidence, subjects were also censored upon reaching their 80^th^ birthday.


```{r, echo=F}
exposure.lag <- 21
employment_status.lag <- 0
hwse3.additional.lag <- 0
hwse2.additional.lag <- 0
spline_yin <- T
spline_year <- T
time_scale <- "age"
hwse2.time_scale <- "age"
table.engine <- "xtable"
messy_sol <- 0.05

```

\fancyhead[R]{Exposure lagged `r exposure.lag` years}

```{r setup, include=F}
knitr::opts_chunk$set(echo = F,
											warning = F,
											message = F,
											cache = F,
											fig.align = 'center',
											fig.pos = 'H',
											results = 'asis')

library(here)
library(pander)
library(knitr)
library(tidyverse)
library(data.table)
library(lubridate)
library(Hmisc)
library(tikzDevice)

source("~/HeadRs/00-my-theme.R")
incidence.key <- fread(here::here("resources", 'cancer-key-expanded.tsv'))
nevent <- readRDS(here::here("resources", "nevent.rds"))
cancers.which <- c(
	# 32:33,
	3, 4, 6:8, 10:12, 15, 18, 19, 21, 25, 31)

```

```{r, eval=F}
if (!('cohort_analytic' %in% ls())) {
	source(here::here('incidence.R'))
}

```

```{r get.tab1, eval=F}
source(here::here('../gm-wrangling/wrangling', 'table1.R'))
tab1 <- get.tab1(
	df = as.data.table(as.data.frame(cohort_analytic[(year >= 1985 | (year >= 1973 & plant %in% 1:2)) & immortal == 0 & right.censored == 0])),
	exposure_lag = exposure.lag,
	use_finrace = T,
	incidence = T)

saveRDS(tab1,
				here::here("./reports/resources/",
									 paste0("tab1", "_lag", exposure.lag, '.rds')))
```

\begin{table}[H]
\caption{Summary of population characteristics. Follow-up for cancer incidence extends from 1985 through 2015.}
\label{tab:tab1}
\centering
```{r tab1}
source(here::here('../gm-wrangling/wrangling', 'table1.R'))
tab1 <- readRDS(
	here::here("reports/resources", paste0("tab1", "_lag", exposure.lag, '.rds')))

tab1 <- tab1[, 1:3]
tab1 <- cbind(rownames(tab1), tab1)

colnames(tab1)[1:4] <- c(" ", "$n$", "$p$", "  ")

description.width <- 8
column.width <- 1.75
table.break <- "Years of follow-"

if (table.engine == 'xtable') {
	tab1 %>% as.data.frame(check.names = F) %>% xtable(
		label = "tab:tab1",
		align = paste0('cp{', description.width, 'cm}',
									 paste0(rep(
									 	paste0('R{', column.width, 'cm}'), ncol(.) - 1
									 ),
									 collapse = ''))
	) %>% print.xtable(
		comment = F,
		add.to.row = list(
			pos = list(grep(table.break, rownames(tab1)) - 1, nrow(.)),
			command = c(
				paste0('\\hline ', '& Median & 25\\textsuperscript{th} \\%tile & 75\\textsuperscript{th} \\%tile \\\\ \n'),
				paste0(
					'\\hline ',
					'\\multicolumn{',
					ncol(.),
					'}{p{',
					description.width + ncol(tab1) * column.width + 1,
					'cm}}{\\footnotesize{',
					'$^\\natural$ Some individuals worked at several sites; plant indicates the site of longest work record time.',
					'}}\\\\',
					'\\multicolumn{',
					ncol(.),
					'}{p{',
					description.width + ncol(tab1) * column.width + 1,
					'cm}}{\\footnotesize{',
					'$^*$ Among those with known date of worker exit.',
					'}}\\\\',
					'\\multicolumn{',
					ncol(.),
					'}{p{',
					description.width + ncol(tab1) * column.width + 1,
					'cm}}{\\footnotesize{',
					'$^\\sharp$ Summary statistics calculated for ever-exposed individuals at end of follow-up only. Exposures were lagged ', exposure.lag, ' years.',
					'}}\\\\'
				)
			)
		)
	)
} else {
	# Remove/replace LaTeX commands
	rownames(tab1) <- gsub(
		"\\\\hspace\\{.*\\}", "&#9;", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\\\hline ", "", rownames(tab1))
	# Footnotes
	# rownames(tab1) <- gsub(
	# 	"\\$\\^\\\\flat\\$", "$^1$", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\$\\^\\\\sharp\\$", "$^3$", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\$\\^\\*\\$", "$^2$", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\$\\^\\\\natural\\$", "$^1$", rownames(tab1))
	
	tab1 <- rbind(
		tab1[1:(which(grepl(table.break, rownames(tab1))) - 1),],
		c("Median", "Q1", "Q3"),
		tab1[which(grepl(table.break, rownames(tab1))):nrow(tab1),]
	)
	
	pander(tab1,
				 justify = c('left', 'right', 'right', 'right'),
				 emphasize.rownames = F,
				 missing = "&nbsp;",
				 caption = tab1.cap
	) %>% cat
	
	paste0(
		'^1^ For individuals who worked at several plants, plant was taken to be the site where they accrued the most work record time.\n\n',
		'^2^ Among those with known date of worker exit.\n\n',
		'^3^ Summary statistics calculated for exposed individuals at end of follow-up only. Exposures were lagged ', exposure.lag, ' years\n\n') %>% cat
}

```
\end{table}

```{r clean_table}
clean.coef.tab <- function(x,
													 table.engine = "xtable",
													 additional.lag = 0) {
	tmp.tab <- as.data.table(as.data.frame(x))
	if (table.engine == 'xtable') {
		tmp.tab[is.na(`(95% CI)`), `(95% CI)` := "\\multicolumn{1}{c}{--}"]
		names(tmp.tab) <- c(
			"Covariate", "level", "$n$", "HR", "(95\\% CI)", "$p$", 'events')
	} else {
		tmp.tab[is.na(`(95% CI)`), `(95% CI)` := "$-$"]
	}

	tmp.tab[!is.na(as.numeric(`$p$`)),`:=`(
		` ` = ifelse(`$p$` < 0.05 & !is.na(as.numeric(`$p$`)), "$*$", "")
	)]

	if (table.engine == 'xtable') {
		tmp.tab[!grepl("0.00|^0$", `$p$`) & !is.na(`$p$`), `:=`(
			`$p$` = paste0("", `$p$`, "\\phantom{0}"))]}
	tmp.tab$`$p$` <- gsub("0.00|^0$", "< 0.005", tmp.tab$`$p$`)

	tmp.tab$`$n$` <- as.character(tmp.tab$`$n$`)
	
	return(tmp.tab)
}

```


\begin{landscape}

\begin{figure}[H]
\caption{Adjusted hazard ratios associated with exposure to straight metal working fluids in the \textbf{UAW-GM Cohort}.}
\label{fig:str}
\centering
\includegraphics[width=7.95in]{"`r here::here(paste0("reports/resources/Lag ", exposure.lag), paste0("str", ifelse(is.finite(messy_sol),
						 paste0("_sol", messy_sol %/% .01), ""), "_facet.pdf"))`"}
\end{figure}

\begin{figure}[H]
\caption{Adjusted hazard ratios associated with exposure to soluble metal working fluids in the \textbf{UAW-GM Cohort}.}
\label{fig:sol}
\centering
\includegraphics[width=7.95in]{"`r here::here(paste0("reports/resources/Lag ", exposure.lag), paste0("sol", ifelse(is.finite(messy_sol),
						 paste0("_sol", messy_sol %/% .01), ""), "_facet.pdf"))`"}
\end{figure}

\begin{figure}[H]
\caption{Adjusted hazard ratios associated with exposure to synthetic metal working fluids in the \textbf{UAW-GM Cohort}.}
\label{fig:syn}
\centering
\includegraphics[width=7.95in]{"`r here::here(paste0("reports/resources/Lag ", exposure.lag),  paste0("syn", ifelse(is.finite(messy_sol),
						 paste0("_sol", messy_sol %/% .01), ""), "_facet.pdf"))`"}
\end{figure}

\begin{figure}[H]
\caption{Adjusted hazard ratios for cancer incidence associated with leaving work in the \textbf{UAW-GM Cohort} (HWSE Condition 2).}
\label{fig:hwse}
\centering
\includegraphics{"`r here::here(paste0("reports/resources/hwse 2/Lag ", exposure.lag + hwse2.additional.lag, ifelse(employment_status.lag == 0, "", paste0("/Employment status lagged ", employment_status.lag, " years"))), paste0("hwse", paste0("_sol", messy_sol %/% .01), "_facet.pdf"))`"}
\end{figure}

\end{landscape}

```{r hwse2}
hwse2.directory <- gsub("//", "/", here::here(paste(
	"./reports/resources/",
	paste0("hwse 2/Lag ", exposure.lag,
				 ifelse(employment_status.lag != 0,
				 			 paste0("/Employment status lagged ", employment_status.lag, " years"), "")),
	ifelse(grepl("age", hwse2.time_scale), "indexed by age", "indexed by calendar"),
	sep = "/"
)))

rbindlist(
	lapply(cancers.which, function(i = cancers.which[1]) {
		outcome <- unlist(incidence.key[i,])
		# Exposure-incidence model
		code <- outcome[1]; description <- outcome[2]
		
		file.prefix <- paste0(
	code,
	ifelse(is.finite(messy_sol),
				 paste0("_sol", messy_sol %/% .01), ""),
	ifelse(spline_year,
				 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
				 ifelse(spline_yin, paste0("_splinedyin"), ""))
)
		
		# HWSE 2 model
		tab.names <- c("hwse2.coef.tab",
									 "hwse2_50.coef.tab",
									 "hwse2_55.coef.tab",
									 "hwse2_60.coef.tab"
									 )
		dir.names <- c("Binary", paste("Age", seq(50, 60, 5))
									 )
		invisible(
		sapply(1:length(tab.names), function (i = 1) {
			assign(tab.names[i],
						 readRDS(
						 	paste0(hwse2.directory, "/",
						 				 dir.names[i], "/",
						 		code,
						 		ifelse(is.finite(messy_sol),
				 paste0("_sol", messy_sol %/% .01), ""),
						 		ifelse(spline_year,
						 					 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
						 					 ifelse(spline_yin, paste0("_splinedyin"))),
							 			 ".tab.rds")), envir = .GlobalEnv)
		}
		))
		
		coef.tab <- rbindlist(list(
			clean.coef.tab(hwse2.coef.tab[1:2,]),
			clean.coef.tab(hwse2_50.coef.tab[1:3,]),
			clean.coef.tab(hwse2_55.coef.tab[1:3,]),
			clean.coef.tab(hwse2_60.coef.tab[1:3,])
		))
		
		# Clean up covariate name
		coef.tab[Covariate == "Cumulative soluble 5",
						 Covariate := "Cumulative soluble"]
		
		coef.tab[grepl("Employment status", Covariate), `:=`(
						 Covariate = c("Employment status", rep(
						 	# "\\cline{2-7}"
						 	NA, .N - 1))
						 )]
		
		# Number of digits
		coef.tab[nchar(gsub("\\$", "", HR)) > 4, `:=`(
			HR = paste0("$", format(as.numeric(gsub("\\$", "", HR)), scientific = T, digits = 3), "$"),
			`(95\\% CI)` = "\\multicolumn{1}{c}{--}"
			# 	paste0(
			# 	"($",
			# 	format(as.numeric(substring(
			# 		`(95\\% CI)`,
			# 		unlist(gregexpr("\\(", `(95\\% CI)`)) + 1,
			# 		unlist(gregexpr("\\,\\\\", `(95\\% CI)`)) - 1)),
			# 		scientific = T, digits = 3), ",\\,",
			# 	format(as.numeric(substring(
			# 	`(95\\% CI)`,
			# 	unlist(gregexpr(",\\\\,", `(95\\% CI)`)) + 3,
			# 	unlist(gregexpr("\\)", `(95\\% CI)`)) - 1)),
			# 	scientific = T, digits = 3),
			# 	")$"
			# )
		)]
		
		# Remove duplicate rows
		coef.tab[duplicated(level), (names(coef.tab)[-1]):=(
			lapply(names(coef.tab)[-1], function(x) {
				NA
			}))]
		
		names(coef.tab)[1] <- "Outcome"
		coef.tab$Outcome <- NA
		
		rbindlist(list(
			data.table(
			Outcome = paste0(ifelse(i != cancers.which[1], "\\\\\n", ""),
			"\\multicolumn{6}{l}{",
			description, " (", coef.tab$events[1], " events)",
			"}\\\\\n%")),
			coef.tab
			), fill = T)
		
	})
) -> coef.tab

names(coef.tab)[1:2] <- c(" ", "  ")

if (table.engine == "xtable") {
			cat("\n\\pagebreak")
			coef.tab[,-'events'] %>% xtable(
				align = c("c", "l", "l", rep("r", ncol(.) - 3), "c"),
				caption = paste0("Adjusted HR estimates for cancer incidence and employment status",
												 ifelse(employment_status.lag == 0, ".", paste0("lagged by ",
												 employment_status.lag, " years.")))
			) %>% print.xtable(
				tabular.environment = "longtable")
		}

```

\newpage

```{r hwse3}
directory <- gsub("//", "/", here::here(paste(
	"./reports/resources/hwse 3",
	paste0("Lag ", exposure.lag + hwse3.additional.lag),
	ifelse(grepl("age", time_scale), "indexed by age", "indexed by calendar"),
	sep = "/"
)))

coef.tab <- readRDS(
	paste0(directory, "/",
		gsub("^_", "", paste0(
			ifelse(is.finite(messy_sol),
						 paste0("_sol", messy_sol %/% .01), ""),
			ifelse(spline_year,
						 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
						 ifelse(spline_yin, paste0("_splinedyin"), ""))
		, ".tab.rds"))))

		coef.tab <- clean.coef.tab(coef.tab)
		
		# Clean up covariate name
		coef.tab[Covariate == "Cumulative soluble 5",
						 Covariate := "Cumulative soluble"]
		
		coef.tab[grepl("Employment status", Covariate), `:=`(
						 Covariate = c("Employment status", rep("\\cline{2-7}", .N - 1))
						 )]
		
		if (table.engine == "xtable") {
			cat("\n\\pagebreak")
			coef.tab[,-'events'] %>% xtable(
				align = c("c", "l", "l", rep("r", ncol(.) - 3), "c"),
				caption = paste0("Adjusted HR estimates for \\textbf{leaving work}.")
			) %>% print.xtable(
				tabular.environment = "longtable")
		}
		

```

```{r coxph}
directory <- gsub("//", "/", here::here(paste(
	"./reports/resources/",
	paste0("Lag ", exposure.lag),
	ifelse(grepl("age", time_scale), "indexed by age", "indexed by calendar"),
	sep = "/"
)))

invisible(
	sapply(cancers.which, function(i = cancers.which[1]) {
		outcome <- unlist(incidence.key[i,])
		# Exposure-incidence model
		code <- outcome[1]; description <- outcome[2]
		
		file.prefix <- paste0(
	code,
	ifelse(is.finite(messy_sol),
				 paste0("_sol", messy_sol %/% .01), ""),
	ifelse(spline_year,
				 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
				 ifelse(spline_yin, paste0("_splinedyin"), ""))
)
		
		coef.tab <- readRDS(
				paste0(directory, "/",
				code, "_sol5",
							 ifelse(spline_year,
							 			 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
							 			 ifelse(spline_yin, paste0("_splinedyin"), "")),
							 ".tab.rds"))
		# HWSE 2 model
		tab.names <- c("hwse2.coef.tab",
									 "hwse2_50.coef.tab",
									 "hwse2_55.coef.tab",
									 "hwse2_60.coef.tab"
									 )
		dir.names <- c("Binary", paste("Age", seq(50, 60, 5))
									 )
		# invisible(
		# sapply(1:length(tab.names), function (i = 1) {
		# 	assign(tab.names[i],
		# 				 readRDS(
		# 				 	paste0(hwse2.directory, "/",
		# 				 				 dir.names[i], "/",
		# 				 		code,
		# 				 		ifelse(is.finite(messy_sol),
		# 		 paste0("_sol", messy_sol %/% .01), ""),
		# 				 		ifelse(spline_year,
		# 				 					 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
		# 				 					 ifelse(spline_yin, paste0("_splinedyin"))),
		# 					 			 ".tab.rds")), envir = .GlobalEnv)
		# }
		# ))
		
		coef.tab <- rbindlist(list(
			clean.coef.tab(coef.tab)
			# clean.coef.tab(hwse2.coef.tab[1:2,]),
			# clean.coef.tab(hwse2_50.coef.tab[1:3,]),
			# clean.coef.tab(hwse2_55.coef.tab[1:3,]),
			# clean.coef.tab(hwse2_60.coef.tab[1:3,])
		))
		
		# Clean up covariate name
		coef.tab[Covariate == "Cumulative soluble 5",
						 Covariate := "Cumulative soluble"]
		
		coef.tab[grepl("Employment status", Covariate), `:=`(
						 Covariate = c("Employment status", rep("\\cline{2-7}", .N - 1))
						 )]
		
		# Number of digits
		coef.tab[nchar(gsub("\\$", "", HR)) > 4, `:=`(
			HR = paste0("$", format(as.numeric(gsub("\\$", "", HR)), scientific = T, digits = 3), "$"),
			`(95\\% CI)` = "\\multicolumn{1}{c}{--}"
			# 	paste0(
			# 	"($",
			# 	format(as.numeric(substring(
			# 		`(95\\% CI)`,
			# 		unlist(gregexpr("\\(", `(95\\% CI)`)) + 1,
			# 		unlist(gregexpr("\\,\\\\", `(95\\% CI)`)) - 1)),
			# 		scientific = T, digits = 3), ",\\,",
			# 	format(as.numeric(substring(
			# 	`(95\\% CI)`,
			# 	unlist(gregexpr(",\\\\,", `(95\\% CI)`)) + 3,
			# 	unlist(gregexpr("\\)", `(95\\% CI)`)) - 1)),
			# 	scientific = T, digits = 3),
			# 	")$"
			# )
		)]
		
		if (table.engine == "xtable") {
			cat("\n\\pagebreak")
			coef.tab[,-'events'] %>% xtable(
				align = c("c", "l", "l", rep("r", ncol(.) - 3), "c"),
				caption = paste0("Adjusted HR estimates for ",
												 ifelse(i <= 31, "incidence of", "death due to"),
												 " \\textbf{", tolower(description), "} ($n = ", unique(coef.tab$events)[1], "$).")
			) %>% print.xtable(
				# hline.after = c(-1, 0,
				# 								nrow(.) - c(11),#,9,6,3),
				# 								nrow(.)),
				tabular.environment = "longtable")
		}
		
	})
)

```

____