---
title: "HWSE path analysis"
subtitle: "Stomach cancer, bladder cancer, and non-Hodgkin Lymphoma"
# author: "Kevin Chen"
date: \today
fontsize: 10pt
output:
  pdf_document:
    latex_engine: lualatex
    includes:
      in_header: "~/HeadRs/StatHead.sty"
    number_sections: false
    toc: false
  keep_tex: true
  # word_document:
  #   reference_docx: "~/HeadRs/style-guide.docx"
  # beamer_presentation:
  #   toc: false
  #   slide_level: 2
  #   # latex_engine: lualatex
  #   includes:
  #     in_header: "~/HeadRs/BeamerHead.sty"
bibliography: "`r path.expand('~/eisen/gm/references/gm.bib')`"
csl: "`r path.expand('~/HeadRs/csl/ama.csl')`"
geometry: margin=2.54cm
---

\onehalfspacing
<!-- \renewcommand{\arraystretch}{1.1} -->
\fancyhead[R]{}
\fancyhead[L]{}
\renewcommand{\headrulewidth}{0pt}

\begin{figure}[H]\begin{center}
\fbox{
\begin{tikzpicture}[>= stealth, auto,
node distance = 3cm, semithick, scale=0.85,
thick, fill=white, inner sep=5pt]
\tikzstyle{every state}=[
shape = rectangle, align = center, draw = none
]
\node[state] (employment) {Employment\\Status};
\node[state] (E1) [below left of=employment, shift=({250:1.3cm})] {Exposure$_{t - 1}$};
\node[state] (E) [below right of=employment]{Exposure$_{t}$};
\node[state] (D) [right of=E1, node distance=9cm] {Disease};
\node[state] (U) [above right of=E] {$U$};
\path[->] (U) edge (employment);
\path[->] (E1) edge (employment);
\path[->] (U) edge (D);
\path[->] (employment) edge (E);
\path[->] (E1) edge (E);
\path[->] (E) edge (D);
\path[->] (E1) edge (D);
\end{tikzpicture}
}
\end{center}
\end{figure}

The presence of the healthy worker survivor effect (HWSE) implies the presence of the following three conditions:

1. Leaving work predicts (future) exposure
2. Leaving work is associated with the disease
3. Prior exposure predicts predicts leaving work

In occupational cohort studies investigating occupational exposure, the first condition readily satisfied. To assess the presence of conditions 2 and 3, we fit Cox proportional hazards models. Associations between prior exposure and disease are also presented here. In particular, the outcomes under consideration are incidences of stomach cancer, bladder cancer, and non-Hodgkin lymphoma.

\textbf{Data.} The analytic data constitute a subset of the UAW-GM Cohort. Person-time is accrued starting in 1941, or three years after hire, whichever came later. Cancer incidence follow-up begins in 1985 for plant 3 and 1973 for plants 1 and 2. Individuals were considered lost to follow-up if they were still alive and cancer-free at age 108 or greater (the maximum observed age at death). For now, those of unknown race were considered to be white.

\textbf{Metalworking fluid exposure and cancer incidence.}

- Person-time starts in 1985 for plant 3 and 1973 for plants 1 and 2 (since we have SEER data for plants 1 and 2).
- Follow-up ends at time of cancer incidence, death, loss to follow-up, or December 31, 2015, whichever comes earlier.
- Cumulative exposure to the three metal working fluid types lagged 21 years are the main exposures of interest.
	- Exposure was categorized into 4 levels including the referent group where possible, and 3 or 2 levels otherwise.
	- The referent group for straight and synthetic metalworking fluids is $0$ mg/m^3^$\cdot$years
	- that for soluble metalworking fluids is 0--0.05 mg/m^3^$\cdot$years.
- Covariates included in the model are splined calendar year, splined year of hire, race, plant, and sex.
- Risk sets were indexed by age.

\textbf{Leaving work and cancer incidence.}

- As in the metalworking fluid-cancer incidence models, follow-up starts starts in 1985 or 1973, depending on the plant.
- Follow-up ends at time of cancer incidence, death, censoring due to end of employment records, or December 31, 2015, whichever comes earlier.
- The main exposure of interest is employment status with no lag and with a lag of 21 years; the levels are:
	- employed,
	- left work under the age of 50, and
	- left work over the age of 50.
- When employment status _was not_ lagged, those who will still at work on January 1, 1995 were censored on that day
- When employment status _was_ lagged, the censor date was lagged by the same amount.
- Covariates included in the model are cumulative metalworking fluid exposure (lagged 1 and 22 years) splined calendar year, splined yer of hire, race, plant, and sex.
- Risk sets were indexed by age.

\textbf{Prior exposure and leaving work.}

- For consistency, follow-up for these models _also_ begins in 1985 or 1973, depending on the plant.
- The outcome of interest is binary employment status, lagged 21 years.
	- Those who will still at work on January 1, 1995 were censored on January 1, 2016.
- The exposures of interest are cumulative metalworking fluid exposure, lagged 22 years.
- Follow-up ends at time of leaving work, death, censoring due to end of employment records, or December 31, 2015, whichever comes earlier.
- Other covariates included in the model are splined calendar year, splined year of hire, race, plant, and sex.
- Risk sets were indexed by age.

```{r, echo=F}
exposure.lag <- 21
employment_status.lag <- c(0, 21)
hwse3.additional.lag <- 21
hwse2.additional.lag <- c(0, 21)
spline_yin <- T
spline_year <- T
time_scale <- "age"
hwse2.time_scale <- "age"
table.engine <- "xtable"
messy_sol <- 0.05

```

```{r setup, include=F}
knitr::opts_chunk$set(echo = F,
											warning = F,
											message = F,
											cache = F,
											fig.align = 'center',
											fig.pos = 'H',
											results = 'asis')

library(here)
library(tidyverse)
library(data.table)
library(lubridate)
library(Hmisc)
library(tikzDevice)

source("~/HeadRs/00-my-theme.R")
incidence.key <- fread(here::here("resources", 'cancer-key-expanded.tsv'))
nevent <- readRDS(here::here("resources", "nevent.rds"))
cancers.which <- grep("stomach|bladder|hodgkin", incidence.key$description, ignore.case = T)

```

```{r, eval=F}
if (!('cohort_analytic' %in% ls())) {
	source(here::here('incidence.R'))
}

```

```{r get.tab1, eval=F}
source(here::here('../gm-wrangling/wrangling', 'table1.R'))
tab1 <- get.tab1(
	df = as.data.table(as.data.frame(cohort_analytic[(year >= 1985 | (year >= 1973 & plant %in% 1:2)) & immortal == 0 & right.censored == 0])),
	exposure_lag = exposure.lag,
	use_finrace = T,
	incidence = T)

saveRDS(tab1,
				here::here("./reports/resources/",
									 paste0("tab1", "_lag", exposure.lag, '.rds')))
```

\begin{table}[H]
\caption{Summary of population characteristics. Follow-up starts in 1973 for plants 1 and 2, and 1985 for plant 3. Follow-up ends in 2015.}
\label{tab:tab1}
\centering
```{r tab1}
source(here::here('../gm-wrangling/wrangling', 'table1.R'))
tab1 <- readRDS(
	here::here("reports/resources", paste0("tab1", "_lag", exposure.lag, '.rds')))

tab1 <- tab1[, 1:3]
tab1 <- cbind(rownames(tab1), tab1)

colnames(tab1)[1:4] <- c(" ", "$n$", "$p$", "  ")

description.width <- 8
column.width <- 1.75
table.break <- "Years of follow-"

if (table.engine == 'xtable') {
	tab1 %>% as.data.frame(check.names = F) %>% xtable(
		label = "tab:tab1",
		align = paste0('cp{', description.width, 'cm}',
									 paste0(rep(
									 	paste0('R{', column.width, 'cm}'), ncol(.) - 1
									 ),
									 collapse = ''))
	) %>% print(
		comment = F,
		add.to.row = list(
			pos = list(grep(table.break, rownames(tab1)) - 1, nrow(.)),
			command = c(
				paste0('\\hline ', '& Median & 25\\textsuperscript{th} \\%tile & 75\\textsuperscript{th} \\%tile \\\\ \n'),
				paste0(
					'\\hline ',
					'\\multicolumn{',
					ncol(.),
					'}{p{',
					description.width + ncol(tab1) * column.width + 1,
					'cm}}{\\footnotesize{',
					'$^\\natural$ Some individuals worked at several sites; plant indicates the site of longest work record time.',
					'}}\\\\',
					'\\multicolumn{',
					ncol(.),
					'}{p{',
					description.width + ncol(tab1) * column.width + 1,
					'cm}}{\\footnotesize{',
					'$^*$ Among those with known date of worker exit.',
					'}}\\\\',
					'\\multicolumn{',
					ncol(.),
					'}{p{',
					description.width + ncol(tab1) * column.width + 1,
					'cm}}{\\footnotesize{',
					'$^\\sharp$ Summary statistics calculated for ever-exposed individuals at end of follow-up only. Exposures were lagged ', exposure.lag, ' years.',
					'}}\\\\'
				)
			)
		)
	)
} else {
	# Remove/replace LaTeX commands
	rownames(tab1) <- gsub(
		"\\\\hspace\\{.*\\}", "&#9;", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\\\hline ", "", rownames(tab1))
	# Footnotes
	# rownames(tab1) <- gsub(
	# 	"\\$\\^\\\\flat\\$", "$^1$", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\$\\^\\\\sharp\\$", "$^3$", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\$\\^\\*\\$", "$^2$", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\$\\^\\\\natural\\$", "$^1$", rownames(tab1))
	
	tab1 <- rbind(
		tab1[1:(which(grepl(table.break, rownames(tab1))) - 1),],
		c("Median", "Q1", "Q3"),
		tab1[which(grepl(table.break, rownames(tab1))):nrow(tab1),]
	)
	
	pander(tab1,
				 justify = c('left', 'right', 'right', 'right'),
				 emphasize.rownames = F,
				 missing = "&nbsp;",
				 caption = tab1.cap
	) %>% cat
	
	paste0(
		'^1^ For individuals who worked at several plants, plant was taken to be the site where they accrued the most work record time.\n\n',
		'^2^ Among those with known date of worker exit.\n\n',
		'^3^ Summary statistics calculated for exposed individuals at end of follow-up only. Exposures were lagged ', exposure.lag, ' years\n\n') %>% cat
}

```
\end{table}

```{r clean_table}
clean.coef.tab <- function(x,
													 table.engine = "xtable",
													 additional.lag = 0) {
	tmp.tab <- as.data.table(as.data.frame(x))
	if (table.engine == 'xtable') { 
		tmp.tab[is.na(`(95% CI)`), `(95% CI)` := "\\multicolumn{1}{c}{--}"]
		names(tmp.tab) <- c(
			"Covariate", "level", "$n$", "HR", "(95\\% CI)", "$p$", 'events')
	} else {
		tmp.tab[is.na(`(95% CI)`), `(95% CI)` := "$-$"]
	}

	tmp.tab[!is.na(as.numeric(`$p$`)),`:=`(
		` ` = ifelse(`$p$` < 0.05 & !is.na(as.numeric(`$p$`)), "$*$", "")
	)]

	if (table.engine == 'xtable') {
		tmp.tab[!grepl("0.00|^0$", `$p$`) & !is.na(`$p$`), `:=`(
			`$p$` = paste0("", `$p$`, "\\phantom{0}"))]}
	tmp.tab$`$p$` <- gsub("0.00|^0$", "< 0.005", tmp.tab$`$p$`)

	tmp.tab$`$n$` <- as.character(tmp.tab$`$n$`)
	
	return(tmp.tab)
}

```


```{r hwse2}
for (j in 1:2) {
hwse2.directory <- gsub("//", "/", here::here(paste(
	"./reports/resources/",
	paste0("hwse 2/Lag ", 1 + hwse2.additional.lag[j],
				 ifelse(employment_status.lag[j] != 0,
				 			 paste0("/Employment status lagged ", employment_status.lag[j], " years"), "")),
	ifelse(grepl("age", hwse2.time_scale), "indexed by age", "indexed by calendar"),
	sep = "/"
)))

rbindlist(
	lapply(cancers.which, function(i = cancers.which[1]) {
		outcome <- unlist(incidence.key[i,])
		# Exposure-incidence model
		code <- outcome[1]; description <- outcome[2]
		
		file.prefix <- paste0(
	code,
	ifelse(is.finite(messy_sol),
				 paste0("_sol", messy_sol %/% .01), ""),
	ifelse(spline_year,
				 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
				 ifelse(spline_yin, paste0("_splinedyin"), ""))
)
		
		# HWSE 2 model
		tab.names <- c("hwse2.coef.tab",
									 "hwse2_50.coef.tab",
									 "hwse2_55.coef.tab",
									 "hwse2_60.coef.tab"
									 )
		dir.names <- c("Binary", paste("Age", seq(50, 60, 5))
									 )
		invisible(
		sapply(1:length(tab.names), function (i = 1) {
			assign(tab.names[i],
						 readRDS(
						 	paste0(hwse2.directory, "/",
						 				 dir.names[i], "/",
						 		code,
						 		ifelse(is.finite(messy_sol),
				 paste0("_sol", messy_sol %/% .01), ""),
						 		ifelse(spline_year,
						 					 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
						 					 ifelse(spline_yin, paste0("_splinedyin"))),
							 			 ".tab.rds")), envir = .GlobalEnv)
		}
		))
		
		coef.tab <- rbindlist(list(
			clean.coef.tab(hwse2.coef.tab[1:2,]),
			clean.coef.tab(hwse2_50.coef.tab[1:3,]),
			clean.coef.tab(hwse2_55.coef.tab[1:3,]),
			clean.coef.tab(hwse2_60.coef.tab[1:3,])
		))
		
		# Clean up covariate name
		coef.tab[Covariate == "Cumulative soluble 5",
						 Covariate := "Cumulative soluble"]
		
		coef.tab[grepl("Employment status", Covariate), `:=`(
						 Covariate = c("Employment status", rep(
						 	# "\\cline{2-7}"
						 	NA, .N - 1))
						 )]
		
		# Number of digits
		coef.tab[nchar(gsub("\\$", "", HR)) > 4, `:=`(
			HR = paste0("$", format(as.numeric(gsub("\\$", "", HR)), scientific = T, digits = 3), "$"),
			`(95\\% CI)` = "\\multicolumn{1}{c}{--}"
			# 	paste0(
			# 	"($",
			# 	format(as.numeric(substring(
			# 		`(95\\% CI)`,
			# 		unlist(gregexpr("\\(", `(95\\% CI)`)) + 1,
			# 		unlist(gregexpr("\\,\\\\", `(95\\% CI)`)) - 1)),
			# 		scientific = T, digits = 3), ",\\,",
			# 	format(as.numeric(substring(
			# 	`(95\\% CI)`,
			# 	unlist(gregexpr(",\\\\,", `(95\\% CI)`)) + 3,
			# 	unlist(gregexpr("\\)", `(95\\% CI)`)) - 1)),
			# 	scientific = T, digits = 3),
			# 	")$"
			# )
		)]
		
		# Remove duplicate rows
		coef.tab[duplicated(level), (names(coef.tab)[-1]):=(
			lapply(names(coef.tab)[-1], function(x) {
				NA
			}))]
		
		names(coef.tab)[1] <- "Outcome"
		coef.tab$Outcome <- NA
		
		rbindlist(list(
			data.table(
			Outcome = paste0(ifelse(i != cancers.which[1], "\\\\\n", ""),
			"\\multicolumn{6}{l}{",
			description, " (", coef.tab$events[1], " events)",
			"}\\\\\n%")),
			coef.tab
			), fill = T)
		
	})
) -> coef.tab

names(coef.tab)[1:2] <- c(" ", "  ")

if (table.engine == "xtable") {
			cat("\n\\pagebreak")
			coef.tab[,-'events'] %>% xtable(
				align = c("c", "l", "l", rep("r", ncol(.) - 3), "c"),
				caption = paste0("Adjusted HR estimates for cancer incidence and employment status",
												 ifelse(employment_status.lag[j] == 0, ".", paste0(" lagged by ",
												 employment_status.lag[j], " years.")))
			) %>% print.xtable(
				tabular.environment = "longtable")
}
}

```

\newpage

```{r hwse3}
directory <- gsub("//", "/", here::here(paste(
	"./reports/resources/hwse 3",
	paste0("Lag ", 1 + hwse3.additional.lag),
	ifelse(employment_status.lag[2] == 0, "",
								paste0("employment status lagged ", employment_status.lag[2], " years")),
	ifelse(grepl("age", time_scale), "indexed by age", "indexed by calendar"),
	sep = "/"
)))

coef.tab <- readRDS(
	paste0(directory, "/",
		gsub("^_", "", paste0(
			ifelse(is.finite(messy_sol),
						 paste0("_sol", messy_sol %/% .01), ""),
			ifelse(spline_year,
						 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
						 ifelse(spline_yin, paste0("_splinedyin"), ""))
		, ".tab.rds"))))

		coef.tab <- clean.coef.tab(coef.tab)
		
		# Clean up covariate name
		coef.tab[Covariate == "Cumulative soluble 5",
						 Covariate := "Cumulative soluble"]
		
		coef.tab[grepl("Employment status", Covariate), `:=`(
						 Covariate = c("Employment status", rep("\\cline{2-7}", .N - 1))
						 )]
		
		if (table.engine == "xtable") {
			cat("\n\\pagebreak")
			coef.tab[,-'events'] %>% xtable(
				align = c("c", "l", "l", rep("r", ncol(.) - 3), "c"),
				caption = paste0("Adjusted HR estimates for \\textbf{leaving work}. ",
																					ifelse(employment_status.lag[2] == 0, "",
																												paste0("Leaving work was lagged ", employment_status.lag[2], " years.")))
			) %>% print.xtable(
				tabular.environment = "longtable")
		}
		

```

```{r coxph}
directory <- gsub("//", "/", here::here(paste(
	"./reports/resources/",
	paste0("Lag ", exposure.lag),
	ifelse(grepl("age", time_scale), "indexed by age", "indexed by calendar"),
	sep = "/"
)))

race_coef.tab <- data.table()

invisible(
	sapply(cancers.which, function(i = cancers.which[1]) {
		outcome <- unlist(incidence.key[i,])
		# Exposure-incidence model
		code <- outcome[1]; description <- outcome[2]
		
		file.prefix <- paste0(
	code,
	ifelse(is.finite(messy_sol),
				 paste0("_sol", messy_sol %/% .01), ""),
	ifelse(spline_year,
				 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
				 ifelse(spline_yin, paste0("_splinedyin"), ""))
)
		
		coef.tab <- readRDS(
				paste0(directory, "/",
				code, "_sol5",
							 ifelse(spline_year,
							 			 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
							 			 ifelse(spline_yin, paste0("_splinedyin"), "")),
							 ".tab.rds"))
		# HWSE 2 model
		tab.names <- c("hwse2.coef.tab",
									 "hwse2_50.coef.tab",
									 "hwse2_55.coef.tab",
									 "hwse2_60.coef.tab"
									 )
		dir.names <- c("Binary", paste("Age", seq(50, 60, 5))
									 )
		# invisible(
		# sapply(1:length(tab.names), function (i = 1) {
		# 	assign(tab.names[i],
		# 				 readRDS(
		# 				 	paste0(hwse2.directory, "/",
		# 				 				 dir.names[i], "/",
		# 				 		code,
		# 				 		ifelse(is.finite(messy_sol),
		# 		 paste0("_sol", messy_sol %/% .01), ""),
		# 				 		ifelse(spline_year,
		# 				 					 ifelse(spline_yin, paste0("_splined"), "_splinedyear"),
		# 				 					 ifelse(spline_yin, paste0("_splinedyin"))),
		# 					 			 ".tab.rds")), envir = .GlobalEnv)
		# }
		# ))
		
		coef.tab <- rbindlist(list(
			clean.coef.tab(coef.tab)
			# clean.coef.tab(hwse2.coef.tab[1:2,]),
			# clean.coef.tab(hwse2_50.coef.tab[1:3,]),
			# clean.coef.tab(hwse2_55.coef.tab[1:3,]),
			# clean.coef.tab(hwse2_60.coef.tab[1:3,])
		))
		
		# Clean up covariate name
		coef.tab[Covariate == "Cumulative soluble 5",
						 Covariate := "Cumulative soluble"]
		
		coef.tab[grepl("Employment status", Covariate), `:=`(
						 Covariate = c("Employment status", rep("\\cline{2-7}", .N - 1))
						 )]
		
		# Number of digits
		coef.tab[nchar(gsub("\\$", "", HR)) > 4, `:=`(
			HR = paste0("$", format(as.numeric(gsub("\\$", "", HR)), scientific = T, digits = 3), "$"),
			`(95\\% CI)` = "\\multicolumn{1}{c}{--}"
			# 	paste0(
			# 	"($",
			# 	format(as.numeric(substring(
			# 		`(95\\% CI)`,
			# 		unlist(gregexpr("\\(", `(95\\% CI)`)) + 1,
			# 		unlist(gregexpr("\\,\\\\", `(95\\% CI)`)) - 1)),
			# 		scientific = T, digits = 3), ",\\,",
			# 	format(as.numeric(substring(
			# 	`(95\\% CI)`,
			# 	unlist(gregexpr(",\\\\,", `(95\\% CI)`)) + 3,
			# 	unlist(gregexpr("\\)", `(95\\% CI)`)) - 1)),
			# 	scientific = T, digits = 3),
			# 	")$"
			# )
		)]
		
		# Save conditional estimates for race
		race_coef.tab <- coef.tab[zoo::na.locf(Covariate) == "Race"]
		race_coef.tab[,`:=`(Outcome = description)]
		assign("race_coef.tab",
					 rbindlist(list(
			get("race_coef.tab", envir = .GlobalEnv),
			race_coef.tab
			)), envir = .GlobalEnv)
		
		if (table.engine == "xtable") {
			cat("\n\\pagebreak")
			coef.tab[,-'events'] %>% xtable(
				align = c("c", "l", "l", rep("r", ncol(.) - 3), "c"),
				caption = paste0("Adjusted HR estimates for ",
												 ifelse(i <= 31, "incidence of", "death due to"),
												 " \\textbf{", tolower(description), "} ($n = ", unique(coef.tab$events)[1], "$).")
			) %>% print.xtable(
				# hline.after = c(-1, 0,
				# 								nrow(.) - c(11),#,9,6,3),
				# 								nrow(.)),
				tabular.environment = "longtable")
		}
		
	})
)

```
