---
title: "HWSE path analysis"
subtitle: "Replicating (?) the analysis in Garcia et al. (2017)"
# author: "Kevin Chen"
date: \today
fontsize: 10pt
output:
  # pdf_document:
  #   latex_engine: lualatex
  #   includes:
  #     in_header: "~/HeadRs/StatHead.sty"
  #   number_sections: false
  #   toc: false
  # keep_tex: true
  # word_document:
  #   reference_docx: "~/HeadRs/style-guide.docx"
  beamer_presentation:
    toc: false
    # latex_engine: lualatex
    includes:
      in_header: "~/HeadRs/BeamerHead.sty"
bibliography: "`r path.expand('~/Box Sync/papers/EnvEpi.bib')`"
csl: "`r path.expand('~/HeadRs/csl/ama.csl')`"
---


```{r setup, include=F}
knitr::opts_chunk$set(echo = F,
											warning = F,
											message = F,
											cache = F,
											fig.align = 'center',
											fig.pos = 'H',
											results = 'asis')

source("~/headrs/00-my-theme.R")
library(data.table)

```

# From Erika Garcia's paper[@Garcia_2017]

<!-- \renewcommand{\arraystretch}{1.1} -->

\begin{figure}[H]\begin{center}
\fbox{
\begin{tikzpicture}[>= stealth, auto,
node distance = 2.5cm, semithick, scale=0.85,
thick, fill=white, inner sep=5pt]
\tikzstyle{every state}=[
shape = rectangle, align = center, draw = none
]
\node[state] (employment) {Employment\\Status$_t$};
\node[state] (E1) [below left of=employment, shift=({250:1cm})] {Exposure$_{t - 1}$};
\node[state] (E) [below right of=employment]{Exposure$_{t}$};
\node[state] (D) [right of=E1, node distance=7.5cm] {Disease};
\node[state] (U) [above right of=E] {$U$};
\path[->] (U) edge (employment);
\path[->] (E1) edge (employment);
\path[->] (U) edge (D);
\path[->] (employment) edge (E);
\path[->] (E1) edge (E);
\path[->] (E) edge (D);
\path[->] (E1) edge (D);
\end{tikzpicture}
}
\end{center}
\end{figure}

The presence of the healthy worker survivor effect (HWSE) implies the presence of the following three conditions:

1. Leaving work predicts (future) exposure
2. Leaving work is associated with the disease
3. Prior exposure predicts predicts leaving work

# Analytic population

- Restricted to
  - Those alive and under age 75 on January 1, 1985
  - Hired in or after 1938
  - Missing no more than half of their work history
- Leaving work and cancer incidence
	- FU spans 1985--1994 (includes SEER)
- MWF exposure and leaving work
	- FU spans 3 years after hire through 1994, death, or leaving work

# An important change!

- Among those with `yrout09_new < yout15`, use `yrout09_new` as the leaving work variable

# Characteristics

```{r, eval=F}
source("../gm-wrangling/wrangling/table1.R")
hwse2.tab1 <- get.tab1(copy(cohort2), mathmode = F)
cancer_male.tab1 <- get.tab1(copy(cohort2[studyno %in% cohort2_male[canc_first == 1 & (ddiag_first <= yoc | is.na(yoc)),studyno] & (year <= year(yoc) | is.na(yoc))]), mathmode = F)
cancer_pr.tab1 <- get.tab1(copy(cohort2[studyno %in% cohort2_male[canc_pr == 1 & (ddiag_pr <= yoc | is.na(yoc)),studyno] & (year <= year(yoc) | is.na(yoc))]), mathmode = F)
cancer_lu.tab1 <- get.tab1(copy(cohort2[studyno %in% cohort2_male[canc_lu == 1 & (ddiag_lu <= yoc | is.na(yoc)),studyno] & (year <= year(yoc) | is.na(yoc))]), mathmode = F)
cancer_corec.tab1 <- get.tab1(copy(cohort2[studyno %in% cohort2_male[canc_corec == 1 & (ddiag_corec <= yoc | is.na(yoc)),studyno] & (year <= year(yoc) | is.na(yoc))]), mathmode = F)
cancer_female.tab1 <- get.tab1(copy(cohort2[studyno %in% cohort2_female[canc_first == 1 & (ddiag_first <= yoc | is.na(yoc)),studyno] & (year <= year(yoc) | is.na(yoc))]), mathmode = F)
hwse3.tab1 <- get.tab1(copy(cohort2_big[year <= year(jobloss.date) & (year <= year(yoc) | is.na(yoc))]), mathmode = F)

save(list = paste0(c("hwse2", "cancer_male", "cancer_pr", "cancer_lu", "cancer_corec", "cancer_female", "hwse3"), ".tab1"),
     file = here::here("reports/replicate garcia 2017", "tab1.rdata"))
```

```{r}
load(file = here::here("reports/replicate garcia 2017", "tab1.rdata"))
clean_tab <- function(tab1 = hwse2.tab1, column_name) {
  na.which <- is.na(tab1[,1])
  tab1[!na.which, 2] <- paste0(
    "(",
    tab1[!na.which, 2],
    ifelse(!is.na(tab1[!na.which, 3]), paste0(", ", tab1[!na.which, 3]), ""),
    ")")

  tab1 <- data.frame(tab1[,1:2])
  
  names(tab1) <- c(paste0("\\multicolumn{2}{c}{", column_name, "}\n%"), "\n")
  
  return(tab1)
}
```

\footnotesize\centering
```{r}
cbind(clean_tab(hwse2.tab1, "Cancer follow-up"),
      " " = NA,
      clean_tab(hwse3.tab1, "Employment follow-up")
			)[
        1:grep("Deceased", rownames(hwse2.tab1)),] %>% xtable(
          align = "lrlcrl"
        ) %>% print(
          include.rownames = T
        )
```

# Cancers in men

\footnotesize\centering

```{r}
cbind(clean_tab(cancer_male.tab1, "All cancers"),
      " " = NA,
      clean_tab(cancer_pr.tab1, "Prostate"))[
        c(1, 20:21, 2:5,18, 17, 24),] %>% xtable(
          align = "p{0.35\\linewidth}rlcrl"
        ) %>% print(include.rownames = T)
```

# Cancers in men

\footnotesize\centering

```{r}
cbind(clean_tab(cancer_lu.tab1, "Lung"),
      " " = NA,
      clean_tab(cancer_corec.tab1, "Colorectal"))[
        c(1, 20:21, 2:5,18, 17, 24),] %>% xtable(
          align = "p{0.35\\linewidth}rlcrl"
        ) %>% print(include.rownames = T)
```

# Cancers in women

\footnotesize\centering

```{r}
clean_tab(cancer_female.tab1, "All cancers")[
        c(1, 20:21, 2:5,18, 17, 24),] %>% xtable(
          align = "lrl"
        ) %>% print(include.rownames = T)
```


# 2. Leaving work and cancer incidence

- Exposure: Employment status (binary)
- Conditioning set:

\begin{columns}
\begin{column}{0.55\linewidth}
\begin{itemize}\item[]\begin{itemize}
\item Age\newline(index time for Cox model)
\item Cumulative MWF exposure (lagged 1 year)
\item Year of hire\newline(P-spline, $df = 3$)
\item Calendar year\newline(P-spline, $df = 3$)
\end{itemize}\end{itemize}
\end{column}
\begin{column}{0.45\linewidth}
\begin{itemize}\item[]\begin{itemize}
\item Duration of employment
\item Race\newline(unknown as white)
\item Plant
\item Sex
\end{itemize}\end{itemize}
\end{column}
\end{columns}

- If cancer incidence date was the same year as leaving work, it was considered to have occurred after leaving work
- If cancer incidence date was after date of death, it was assumed to have been on the day of death

# 2. Leaving work by age 50 and cancer incidence

$$\begin{aligned}
\log h(t \mid a,\ x)
& = \log h_0 (t) \\ & \hspace{2em}
+ a \cdot \Ind{t < 51} \cdot \beta_1
+ a \cdot \Ind{t \ge 51} \cdot \beta_2 \\ & \hspace{2em}
+ x \begin{pmatrix}\beta_3 & \cdots & \beta_p\end{pmatrix}^\top
\end{aligned}$$

where $a$ is the indicator of having left work, $t$ is age, and $x$ is a vector of covariates

- Coefficients $\beta_1$ and $\beta_2$ may be thought of as interaction effects of employment status and age

# 2. Leaving work and cancer incidence

\centering\includegraphics[width=0.8\linewidth]{`r here::here("reports/resources/replicate garcia 2017/hwse 2/FU through 1994/Lag 1/indexed by age", "hwse2.pdf")`}

# 2. Leaving work and cancer incidence (what I did most recently, but new YOUT and FU until 75)

\centering\includegraphics[width=0.8\linewidth]{`r here::here("reports/resources/garcia yout/under 75/hwse 2/FU 1985 through 1994/Lag 1/indexed by age", "hwse2_selected.pdf")`}

# 2. Leaving work and cancer incidence (lagged employment status)

\centering\includegraphics[width=0.8\linewidth]{`r here::here("reports/resources/garcia yout/under 75/hwse 2/FU through 1994/Lag 1/Employment status lagged 1 years/indexed by age", "hwse2_selected.pdf")`}


# 3. Prior exposure and leaving work

- Exposure: Cumulative exposure lagged 1 year
- Conditioning set:

\begin{columns}
\begin{column}{0.55\linewidth}
\begin{itemize}\item[]\begin{itemize}
\item Age\newline(index time for Cox model)
\item Calendar year\newline(P-spline $df = 3$)
\end{itemize}\end{itemize}
\end{column}
\begin{column}{0.45\linewidth}
\begin{itemize}\item[]\begin{itemize}
\item Race\newline(unknown as white)
\item Plant
\end{itemize}\end{itemize}
\end{column}
\end{columns}

# 3. Prior exposure and leaving work (men)

```{r}
hwse3.dir <- here::here(paste0(
		"./reports/resources/replicate garcia 2017",
		paste0(paste0("/hwse 3",
									paste0("/FU through 1994"),
									"/Lag ", 1),
					 "/indexed by age"
		)))

men_coef.tab <- readRDS(paste0(hwse3.dir, "/sol127_splinedyear.coef.rds"))
women_coef.tab <- readRDS(paste0(hwse3.dir, "/female/sol10_splinedyear.coef.rds"))
```

\footnotesize\centering

```{r}
men_coef.tab[,.(
  `\\multicolumn{2}{l}{MWF exposure}\n%` = c(
    paste0("\\multicolumn{5}{l}{", gsub(" [0-9].*", "", Covariate[1]), "}\\\\\n"),
    rep(NA, .N - 1)),
  "\n" = level, n,
  `Adjusted HR` = round(HR, 2),
  `(95\\% CI)` = ifelse(is.na(lower.ci), "", paste0(
    "(", round(lower.ci, 2), ", ", round(upper.ci, 2), ")"
  ))
), Covariate][grepl("Cumulative", Covariate),-1] %>% xtable

```

# 3. Prior exposure and leaving work (women)

\footnotesize\centering

```{r}
women_coef.tab[,.(
  `\\multicolumn{2}{l}{MWF exposure}\n%` = c(
    paste0("\\multicolumn{5}{l}{", gsub(" [0-9].*", "", Covariate[1]), "}\\\\\n"),
    rep(NA, .N - 1)),
  "\n" = level, n,
  `Adjusted HR` = round(HR, 2),
  `(95\\% CI)` = ifelse(is.na(lower.ci), "", paste0(
    "(", round(lower.ci, 2), ", ", round(upper.ci, 2), ")"
  ))
), Covariate][grepl("Cumulative", Covariate),-1] %>% xtable


```


# 3. Prior exposure and leaving work (what I did most recently)

```{r}
hwse3.dir <- here::here(paste0(
		"./reports/resources",
		paste0(paste0("/hwse 3",
									paste0("/FU through 1994"),
									"/Lag ", 1),
					 "/indexed by age"
		)))

hwse3_coef.tab <- readRDS(paste0(hwse3.dir, "/sol5_splined.coef.rds"))
```

\footnotesize\centering

```{r}
hwse3_coef.tab[,.(
  `\\multicolumn{2}{l}{MWF exposure}\n%` = c(
    paste0("\\multicolumn{5}{l}{", gsub(" [0-9].*", "", Covariate[1]), "}\\\\\n"),
    rep(NA, .N - 1)),
  "\n" = level, n,
  `Adjusted HR` = round(HR, 2),
  `(95\\% CI)` = ifelse(is.na(lower.ci), "", paste0(
    "(", round(lower.ci, 2), ", ", round(upper.ci, 2), ")"
  ))
), Covariate][grepl("Cumulative", Covariate),-1] %>% xtable

```

# 3. Prior exposure and leaving work (what I did most recently, but new YOUT and FU until 75)

```{r}
hwse3.dir <- here::here(paste0(
		"./reports/resources/garcia yout/under 75",
		paste0(paste0("/hwse 3",
									paste0("/FU through 1994"),
									"/Lag ", 1),
					 "/indexed by age"
		)))

hwse3_new_yout_coef.tab <- readRDS(paste0(hwse3.dir, "/sol5.coef.rds"))
```

\footnotesize\centering

```{r}
hwse3_new_yout_coef.tab[,.(
  `\\multicolumn{2}{l}{MWF exposure}\n%` = c(
    paste0("\\multicolumn{5}{l}{", gsub(" [0-9].*", "", Covariate[1]), "}\\\\\n"),
    rep(NA, .N - 1)),
  "\n" = level, n,
  `Adjusted HR` = round(HR, 2),
  `(95\\% CI)` = ifelse(is.na(lower.ci), "", paste0(
    "(", round(lower.ci, 2), ", ", round(upper.ci, 2), ")"
  ))
), Covariate][grepl("Cumulative", Covariate),-1] %>% xtable

```

# Citations
